A Mechanised Library for Generic (Co)Programming

### 2024-02-27 paper/paper.tex
Some style changes

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
\newcommand{\coq}[1]{\mintinline{coq}{#1}}

@@

\mpav{I need this to be written in Coq notation}
\begin{minted}{coq}
qs :: [a] -> [a]
qs [] = []
qs (x:xs) = qs (filter (x <=) xs) ++ [x] ++ (qs (filter (x >) xs))
\end{minted}
This program is formed of essentially two parts. The first one is what is
commonly thought as the \emph{partition} function destructuring the input list
in three parts, the pivot and the other two sublists, and the second part
recursing over the sublists and composing the results back together.

Notice that if we define the functor \coq{TreeF} $X = X \times A \times X$ with
the obvious functorial action then \coq{partition} can be viewed as a
\coq{TreeF}-coalgebra on lists over the type $A$
\begin{minted}{coq}
c : [A] -> TreeF [A]
c (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
\end{minted}

While we can define an \coq{TreeF}-algebra on lists
over $A$ as well, that is a map
\begin{minted}{coq}
a : TreeF [A] -> [A]
a (left, x, right) = left ++ [x] ++ right
\end{minted}

@@

\subsection{Functors and Containers}
\label{sec:containers}
A container is defined by a type of \emph{shapes} and a \emph{family of position
types} indexed by shapes.
\begin{minted}{Coq}
Context (Shape : Type) (Pos : Shape -> Type).
\end{minted}
An \emph{extension} of this container is a functor defined as follows:
\begin{minted}{Coq}
Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
\end{minted}
with the obvious action on morphisms. Generally, any functor that is isomorphic
to a container extension is called a \emph{container functor}. However, in our
work we will call \emph{container functor} to the functors defined by a specific
container extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \coq{Valid} that
determines whether a position is valid in a shape.
\begin{minted}{Coq}
Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
Record App (X : Type) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{minted}
This approach solves the the problem of the unsafe casts, since the code
extracted for \coq{{p | Valid shape pos}} will be an OCaml singleton
type defined as the OCaml equivalent to \coq{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \coq{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{minted}{Coq}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

The Coq value \coq{inl n} of type
\coq{nat + bool * bool} would look as follows as a container
extension 
\coq{App (Sum (Const nat) (Prod Id Id)) bool}:
\begin{minted}{Coq}
{| shape := inl n; cont := fun e => uninhabited e|}
\end{minted}
We omit the definition of \coq{uninhabited}, which exploits the
fact that \coq{e} for \coq{inl n} implies
\coq{False} to build an element of type \coq{bool}%
\footnote{Note that, since \coq{e} is uninhabited we could return
any arbitrary value instead of \coq{uninhabited e}.}.

Similarly, the Coq value \coq{inr (true, false)} as a container
extension is as follows:
\begin{minted}{Coq}
  {| shape := inr (tt,tt);
     cont := fun e : {p | valid (inr (tt, tt)) p} =>
        match proj1_sig e with
        | inl f => match f with end
        | inr (inl tt) => fun _ => true
        | inr (inr tt) => fun _ => false
        end
  |}
\end{minted}
Here, the position \coq{inl f} is impossible, as
\coq{f} would need to be a proof of \coq{False}.
Of the two remaining positions, the left one will return 
\coq{true}, and the right position \coq{false}.
\begin{figure}
\begin{minipage}{.52\textwidth}
  \begin{minted}[linenos, fontsize=\small]{Coq}
Context `(C1 : Cont S1 P1) 
        `(C2 : Cont S2 P2).

@@

\subsubsection{Equality of Container Extensions}
The equality of container extensions needs to be extensional, since the
contents are obtained by applying a function. We want to equate container
extensions that have the same shape, and that, for equal shape and position,
they return the same element. To avoid the use of functional extensionality
axioms, we capture this relation with the following inductive proposition in
Coq:
\begin{minted}{Coq}
Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
| AppR_ext (Es : shape x = shape y)
           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
\end{minted}
Note that  we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is
trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \coq{App F A -> B} (or \coq{B -> App F A}). Since 
these types now come with their own equivalence, any function that manipulates
them needs to be \emph{respectful}. I.e.\ given 
\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{minted}{Coq}
forall (x y : X), R x y -> R' (f x) (f y)
\end{minted}

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of setoids,
i.e.\ types with na associated equivalence relation, and considering only
functions that respect the associated equivalences, i.e. \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation.  We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code.  We use \coq{=e} to denote
the equivalence relation of a setoid. By default, we associate every Coq type
with the standard propositional equality, unless a different equivalence is
specified (i.e.\ we allow overlapping instances, and Coq's propositional
equality takes the lowest priority). 
\begin{minted}{Coq}
Reserved Notation "f =e g" (at level 70, no associativity).
Class setoid A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{minted}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{minted}{Coq}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph
    { app :> A -> B;
      app_eq : forall x y, x =e y -> app x =e app y
    }.
Notation "A ~> B" := (@morph A _ B _).
\end{minted}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \coq{Prop} in the code, so objects of type
\coq{A ~> B} will be extracted to the OCaml equivalent to
\coq{A -> B}.

\subsubsection{Polynomial Types}
We define a number of equivalences for polynomial types. These equivalences are
be used when building proper morphisms that manipulate their respective types.
\begin{minted}{Coq}
Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
Instance pair_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A * B).
Instance sum_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A + B).
Instance prop_eq : setoid Prop.
Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
\end{minted}

@@

Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{minted}{Coq}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape ~> Pos ~> bool }.
Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{minted}
Note the use of \coq{=e} instead of Coq's standard equality.  For
positions, however, we chose to use Coq's propositional equality, since these
would leads to simpler code. We explain why in Section~\ref{sec:cata}, and the
mechanisation of initial algebras of container extensions. The definitions in
Figure~\ref{fig:poly} are redefined to account for these changes in a
straightforward way.

We prove the isomorphisms of polynomial types and the equivalent (redefined)
container extensions. As an example, we discuss (informally) the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \coq{App F X} is isomorphic to \coq{A}, and
\coq{App G X} is isomorphic to \coq{B}. Then we can
show that \coq{App (Prod F G) X} is isomorphic to
\coq{A * B}.  If we have an element of type 
\coq{App (Prod F G) X},
using the \coq{inl} position, we can obtain
\coq{App F X}.  Similarly, using \coq{inr}, we can
obtain \coq{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \coq{App F X} and \coq{App G X} to
obtain \coq{A * B}. Similarly if we have \coq{A * B},
we can first use the isomorphisms of 
\coq{A} and
\coq{B} to obtain
\coq{App F X * App G X}, and then construct the necessary container
extension. Given \coq{p_inl}/\coq{p_inr} that act as
\coq{inl/inr} but on valid product positions, the relevant
functions that define the isomorphism are:
\begin{minted}{Coq}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (e_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (e_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := fun e => match val e with 
                    | inl p => fun H => cont (fst x) (ValidPos p H)
                    | inr p => fun H => cont (snd x) (ValidPos p H)
                    end (IsValid e)
|}.
\end{minted}
Proving that these functions indeed form an isomorphism is straightforward,
using the fact that Product containers only have two valid positions.

@@

Algebras on container extensions are defined as expected. The carrier of the
algebra is \coq{A}, and the operations of the algebra are morphisms
of the following type:
\begin{minted}{Coq}
Notation Alg F A := (App F A ~> A).
\end{minted}
Given a container extension \coq{F}, we define \emph{two}
(isomorphic) initial algebras. We will show the need for the second initial
algebra in the following section. Our fist initial algebra is defined as the
straightforward least fixpoint of the container extension \coq{F}:
\begin{minted}{Coq}
Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
\end{minted}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words,
we would need to provide an equivalence for \coq{LFix}, but the
only way to do this is to define such equivalence simultaneously to the
definition for  \coq{LFix}.  We could have overcame this
circularity by defining  \coq{LFix} independently of container
extensions, and define later the necessary
\coq{in}/\coq{out} functions to convert between
\coq{LFix} and \coq{App}. However, this extra layer
interferes with code extraction, in that it requires destructing/constructing
subsequent \coq{LFix} and  \coq{App}, which Coq is not
always able to inline.

We provide a custom induction scheme for initial algebras:
\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \coq{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \coq{LFixR}.
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, a catamorphism takes an
\coq{F}-algebra with carrier \coq{A}, and produces a
function from \coq{LFix F} to \coq{A}.
\begin{minted}{Coq}
Definition cata_f `{setoid A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkApp sx (fun e => f (kx e)))
  end.
\end{minted}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{minted}{Coq}
Definition cata `{eA : setoid A} : Alg F A ~> LFix ~> A := ...
\end{minted}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \coq{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \coq{f} and
\coq{cata g}.
\begin{minted}{Coq}
Lemma cata_univ `{eA : setoid A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{minted}

@@

Terminal coalgebras are defined analogously to initial algebras:
\begin{minted}{Coq}
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
\end{minted}
Their equivalence relation, and anamorphisms, and universal properties are
easy to encode:
\begin{minted}{Coq}
  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...

>>>>

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}

%\newcommand{\coq}[1]{\mintinline{coq}{#1}}
\newmintinline[coq]{coq}{fontsize=\small}
\newminted{coq}{fontsize=\small}

@@

\mpav{I need this to be written in Coq notation}
\begin{coqcode}
qs :: [a] -> [a]
qs [] = []
qs (x:xs) = qs (filter (x <=) xs) ++ [x] ++ (qs (filter (x >) xs))
\end{coqcode}
This program is formed of essentially two parts. The first one is what is
commonly thought as the \emph{partition} function destructuring the input list
in three parts, the pivot and the other two sublists, and the second part
recursing over the sublists and composing the results back together.

Notice that if we define the functor \coq{TreeF} $X = X \times A \times X$ with
the obvious functorial action then \coq{partition} can be viewed as a
\coq{TreeF}-coalgebra on lists over the type $A$
\begin{coqcode}
c : [A] -> TreeF [A]
c (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
\end{coqcode}

While we can define an \coq{TreeF}-algebra on lists
over $A$ as well, that is a map
\begin{coqcode}
a : TreeF [A] -> [A]
a (left, x, right) = left ++ [x] ++ right
\end{coqcode}

@@

\subsection{Functors and Containers}
\label{sec:containers}
A container is defined by a type of \emph{shapes} and a \emph{family of position
types} indexed by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
An \emph{extension} of this container is a functor defined as follows:
\begin{coqcode}
Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
\end{coqcode}
with the obvious action on morphisms. Generally, any functor that is isomorphic
to a container extension is called a \emph{container functor}. However, in our
work we will call \emph{container functor} to the functors defined by a specific
container extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \coq{Valid} that
determines whether a position is valid in a shape.
\begin{coqcode}
Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
Record App (X : Type) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{coqcode}
This approach solves the the problem of the unsafe casts, since the code
extracted for \coq{{p | Valid shape pos}} will be an OCaml singleton
type defined as the OCaml equivalent to \coq{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \coq{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{coqcode}
Then, we define \emph{container extensions} as
\begin{coqcode}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{coqcode}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{coqcode}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{coqcode}

@@

The Coq value \coq{inl n} of type
\coq{nat + bool * bool} would look as follows as a container
extension 
\coq{App (Sum (Const nat) (Prod Id Id)) bool}:
\begin{coqcode}
{| shape := inl n; cont := fun e => uninhabited e|}
\end{coqcode}
We omit the definition of \coq{uninhabited}, which exploits the
fact that \coq{e} for \coq{inl n} implies
\coq{False} to build an element of type \coq{bool}%
\footnote{Note that, since \coq{e} is uninhabited we could return
any arbitrary value instead of \coq{uninhabited e}.}.

Similarly, the Coq value \coq{inr (true, false)} as a container
extension is as follows:
\begin{coqcode}
  {| shape := inr (tt,tt);
     cont := fun e : {p | valid (inr (tt, tt)) p} =>
        match proj1_sig e with
        | inl f => match f with end
        | inr (inl tt) => fun _ => true
        | inr (inr tt) => fun _ => false
        end
  |}
\end{coqcode}
Here, the position \coq{inl f} is impossible, as
\coq{f} would need to be a proof of \coq{False}.
Of the two remaining positions, the left one will return 
\coq{true}, and the right position \coq{false}.
\begin{figure}
\begin{minipage}{.52\textwidth}
  \begin{minted}[linenos, fontsize=\small]{Coq}
Context `(C1 : Cont S1 P1) 
        `(C2 : Cont S2 P2).

@@

\subsubsection{Equality of Container Extensions}
The equality of container extensions needs to be extensional, since the
contents are obtained by applying a function. We want to equate container
extensions that have the same shape, and that, for equal shape and position,
they return the same element. To avoid the use of functional extensionality
axioms, we capture this relation with the following inductive proposition in
Coq:
\begin{coqcode}
Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
| AppR_ext (Es : shape x = shape y)
           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
\end{coqcode}
Note that  we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is
trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \coq{App F A -> B} (or \coq{B -> App F A}). Since 
these types now come with their own equivalence, any function that manipulates
them needs to be \emph{respectful}. I.e.\ given 
\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{coqcode}
forall (x y : X), R x y -> R' (f x) (f y)
\end{coqcode}

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of setoids,
i.e.\ types with na associated equivalence relation, and considering only
functions that respect the associated equivalences, i.e. \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation.  We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code.  We use \coq{=e} to denote
the equivalence relation of a setoid. By default, we associate every Coq type
with the standard propositional equality, unless a different equivalence is
specified (i.e.\ we allow overlapping instances, and Coq's propositional
equality takes the lowest priority). 
\begin{coqcode}
Reserved Notation "f =e g" (at level 70, no associativity).
Class setoid A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph
    { app :> A -> B;
      app_eq : forall x y, x =e y -> app x =e app y
    }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \coq{Prop} in the code, so objects of type
\coq{A ~> B} will be extracted to the OCaml equivalent to
\coq{A -> B}.

\subsubsection{Polynomial Types}
We define a number of equivalences for polynomial types. These equivalences are
be used when building proper morphisms that manipulate their respective types.
\begin{coqcode}
Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
Instance pair_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A * B).
Instance sum_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A + B).
Instance prop_eq : setoid Prop.
Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
\end{coqcode}

@@

Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape ~> Pos ~> bool }.
Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For
positions, however, we chose to use Coq's propositional equality, since these
would leads to simpler code. We explain why in Section~\ref{sec:cata}, and the
mechanisation of initial algebras of container extensions. The definitions in
Figure~\ref{fig:poly} are redefined to account for these changes in a
straightforward way.

We prove the isomorphisms of polynomial types and the equivalent (redefined)
container extensions. As an example, we discuss (informally) the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \coq{App F X} is isomorphic to \coq{A}, and
\coq{App G X} is isomorphic to \coq{B}. Then we can
show that \coq{App (Prod F G) X} is isomorphic to
\coq{A * B}.  If we have an element of type 
\coq{App (Prod F G) X},
using the \coq{inl} position, we can obtain
\coq{App F X}.  Similarly, using \coq{inr}, we can
obtain \coq{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \coq{App F X} and \coq{App G X} to
obtain \coq{A * B}. Similarly if we have \coq{A * B},
we can first use the isomorphisms of 
\coq{A} and
\coq{B} to obtain
\coq{App F X * App G X}, and then construct the necessary container
extension. Given \coq{p_inl}/\coq{p_inr} that act as
\coq{inl/inr} but on valid product positions, the relevant
functions that define the isomorphism are:
\begin{coqcode}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (e_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (e_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := fun e => match val e with 
                    | inl p => fun H => cont (fst x) (ValidPos p H)
                    | inr p => fun H => cont (snd x) (ValidPos p H)
                    end (IsValid e)
|}.
\end{coqcode}
Proving that these functions indeed form an isomorphism is straightforward,
using the fact that \coq{Prod} containers only have two valid positions.

@@

Algebras on container extensions are defined as expected. The carrier of the
algebra is \coq{A}, and the operations of the algebra are morphisms
of the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
Given a container extension \coq{F}, we define \emph{two}
(isomorphic) initial algebras. We will show the need for the second initial
algebra in the following section. Our fist initial algebra is defined as the
straightforward least fixpoint of the container extension \coq{F}:
\begin{coqcode}
Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
\end{coqcode}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words,
we would need to provide an equivalence for \coq{LFix}, but the
only way to do this is to define such equivalence simultaneously to the
definition for  \coq{LFix}.  We could have overcame this
circularity by defining  \coq{LFix} independently of container
extensions, and define later the necessary
\coq{in}/\coq{out} functions to convert between
\coq{LFix} and \coq{App}. However, this extra layer
interferes with code extraction, in that it requires destructing/constructing
subsequent \coq{LFix} and  \coq{App}, which Coq is not
always able to inline.

We provide a custom induction scheme for initial algebras:
\begin{coqcode}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{coqcode}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \coq{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \coq{LFixR}.
\begin{coqcode}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{coqcode}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, a catamorphism takes an
\coq{F}-algebra with carrier \coq{A}, and produces a
function from \coq{LFix F} to \coq{A}.
\begin{coqcode}
Definition cata_f `{setoid A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkApp sx (fun e => f (kx e)))
  end.
\end{coqcode}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{coqcode}
Definition cata `{eA : setoid A} : Alg F A ~> LFix ~> A := ...
\end{coqcode}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \coq{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \coq{f} and
\coq{cata g}.
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{coqcode}

@@

Terminal coalgebras are defined analogously to initial algebras:
\begin{coqcode}
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}
Their equivalence relation, and anamorphisms, and universal properties are
easy to encode:
\begin{coqcode}
  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...

  Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
This definition, however, prevents us from composing catamorphisms and
anamorphisms. To address this issue, we consider another alternative definition
for initial $F$-algebras. In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata},
and they can be used to define the isomorphism between 
\coq{{x : GFix F | FinF x}} and
\coq{LFix F}. In particular: 
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, anamorphisms produce finite trees,
then their return type can also be the \emph{least} fixed point of the
container extension. We encode in Coq the notion of \emph{recursive
anamorphisms} and \emph{recursive coalgebras}~\cite{TODO}.  Recursive
coalgebras are coalgebras whose anamorphism produces finite trees.  The
following predicate captures this fact. The following predicate states that
\coq{h} can only be applied finitely many times:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
We show how this predicate is equivalent to the anamorphism of such coalgebras
producing finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof that they can only be applied finitely many times
(\coq{RecF}).  Note that such proofs will be erased when extracting
code to OCaml. This is standard practice when building general recursive
functions in Coq\footnote{\dcas{This technique had a name in the community, but
I cannot remember. Something predicates}}. The core definitions are the
following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
